Project 3 - Exploring Prosper Loan Data by Li Liang
========================================================
I chose the Prosper loan data for project 3. I wanted to explore the P2P 
lender's data to find insights related to the borrowers and the loans they 
broker. At quick glance, there are over 100k rows of loans and 81 variables. 
I initally choose way too many variables (30+) and through exploratory 
analysis, whittled it down to 20 variables and even created a few variables. 
Let's begin.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.
library(dplyr)
library(ggplot2)
library(lubridate)
library(corrplot)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

## setting up the working directory

##getwd()
#setwd('c:/users/li/documents')
loans <- read.csv('loans.csv')

## Storing loan dataset into dataframe l
l <- select(loans, ListingCategory..numeric.,
            EmploymentStatusDuration,
            CreditScoreRangeLower,
            CreditScoreRangeUpper,
            CurrentCreditLines,
            DebtToIncomeRatio,
            StatedMonthlyIncome,
            BorrowerState,
            Occupation,
            EmploymentStatus,
            IsBorrowerHomeowner,
            IncomeRange,
            IncomeVerifiable,
            BorrowerAPR,
            BorrowerRate,
            Term,
            LoanStatus,
            LoanOriginalAmount,
            LoanOriginationDate,
            MonthlyLoanPayment
)
l$Term <- factor(l$Term)
l$ListingCategory..numeric. <- factor(l$ListingCategory..numeric.)
cats <- c('0 - Not Available', '1 - Debt Consolidation', 
          '2 - Home Improvement', '3 - Business', '4 - Personal Loan', 
          '5 - Student Use', '6 - Auto', '7- Other', '8 - Baby&Adoption', 
          '9 - Boat', '10 - Cosmetic Procedure', '11 - Engagement Ring', 
          '12 - Green Loans', '13 - Household Expenses', 
          '14 - Large Purchases', '15 - Medical/Dental', '16 - Motorcycle', 
          '17 - RV', '18 - Taxes', '19 - Vacation', '20 - Wedding Loans') 
summary(l)
str(l)
```

# Univariate Plots Section
 This section explores the variables in the data set in form of
univariate charts and plots. Variables such as: Loan amounts, credit scores,
borrower rates, borrower financial stats, loan origination, loan status, loans
by income, loans by month, loan types, dollars loaned, # of loans, 
defaults by year, state, employment status and occupation. 

**Note: I only use CreditScoreRangeLower throughout this analysis to describe
credit scores. CreditScoreRangeHigher has very similar distribution but is 20
points higher.** 

```{r, Loan Amounts, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=LoanOriginalAmount)) +geom_bar(binwidth=700)
```

 Loan amount exhibits positive skew, mean 8300 > median 6500.


```{r, Loan Payments, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=MonthlyLoanPayment)) +geom_bar(binwidth=20) +xlim(0, 1500)
```

Monthly loan payments exhibits positive skew, mean 272 > median 217.

```{r, Credit Scores, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=CreditScoreRangeLower)) +geom_bar(binwidth=5) +
  xlim(400, 880)

```

Credit ranges are normally distributed with mean/med = 685/680. It seem they 
have common intervals and are bunched as such. 
**Note: As noted before only CreditScoreRangeLower is used.** 

```{r, Borrower Interest Rate, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=BorrowerRate)) +geom_bar(binwidth=.01)
```

Interest rate are normally distributed mean/med = .19 However, there is a 
spike around .31.  I'll take a closer look in bivariate section.

```{r, Debt to Income Ratio, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=DebtToIncomeRatio)) +geom_bar() + xlim(0, 1.5)
```

DI ratio exhibits slight positive skew mean .27 > median .22. There are more 
people with DI < 0.5.


```{r, Stated Monthly Income, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=StatedMonthlyIncome)) +geom_bar() +xlim(0, 20000)
```

Monthly income exhibits positive skew mean 5600 > median 4666. 

```{r,   estimated monthly debt, echo=FALSE, message=FALSE , warning=FALSE}
# Creating an estimate of total monthly debt by multiplying DI ratio by 
# stated monthly income

l$TotalMonthlyDebt<- l$DebtToIncomeRatio * l$StatedMonthlyIncome

ggplot(data=subset(l, !is.na(TotalMonthlyDebt)), aes(x=TotalMonthlyDebt)) +
  geom_bar(binwidth=100) + xlim(0, 4000)

summary(l$TotalMonthlyDebt)
```

I created the variable TotalMonthlyDebt to estimate the monthly debt at 
inception of loan by multiplying the DI ratio by the stated monthly income. 
The resulting distribution is positively skewed. 

```{r, Employment Status Duration, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=EmploymentStatusDuration)) +geom_bar(binwidth=1)
```

Employment duration exhibits positive skew mean 96 > median 67. It seems that 
more people that are employed shorter durations.

```{r, Credit Lines, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=CurrentCreditLines)) +geom_bar(binwidth=1)
```

Current credit lines has a slight positive skew and the mean/med = 10.


```{r, Monthly, Years, echo=FALSE, message=FALSE , warning=FALSE}

#Converting date into lubridate type in order to parse out month, year
l$ldate <- ymd_hms(l$LoanOriginationDate)
l$LoanMonth <- factor(month(l$ldate))
l$LoanYear <- factor(year(l$ldate))

ggplot(data=l, aes(x=LoanMonth)) +geom_bar()


ggplot(data=l, aes(x=LoanYear)) +geom_bar()
```

I decomposed loan origination date using lubridate and plotted by month
and year. I observed seasonality in the loans by month by during the
holiday season. It is also interesting in the loans by year plot that the
loans dropped off in 2009. I would guess that is a supply side shortage due
to the Financial Crisis. What do you think? (Hint: I'm wrong.)


```{r, Status, echo=FALSE, message=FALSE , warning=FALSE}

table(l$LoanStatus)

```

Most loans in the whole data set are completed and current. Later I will 
break it down by year. 


```{r,  Tables, echo=FALSE, message=FALSE , warning=FALSE}
print('# of Loans by Income Range')
summary(l$IncomeRange)
print('# of Loans by Term (months)')
summary(l$Term)
```

Income range is mostly in the middle from $25k-75k with mostly 3 year loans. 


```{r,  loan count vs date by term, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=ldate, fill=l$Term)) +geom_bar() +
  scale_fill_discrete(name="Loan Term (Months)") + xlab("")
```

Graph of loan term by year. Dominated by 3 year loans. 


```{r,  Loan Category, echo=FALSE, message=FALSE,fig.width = 8, fig.height = 10}
# plot of count of loan types
ggplot(data=l,aes(x=ListingCategory..numeric., fill=ListingCategory..numeric.))+
  geom_bar() + scale_fill_discrete(name="Loan Category", labels=cats)

# aggregation of $ total loaned by type
Loansbydollar <- l %>%
  group_by(ListingCategory..numeric.) %>%
    summarise(DollarTotal=sum(LoanOriginalAmount), CountTotal=n()) %>%
  arrange(desc(DollarTotal, CountTotal))

#print
Loansbydollar

```

Debt consolidation by far is the leading loan category in number of loans
and $ amount. But it seems there are a lot of loans not classified. 
More on this in the multivarate section.

```{r,  re, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=LoanYear, fill=LoanStatus)) +geom_bar() 
```

Loan status broken down by year you can see more texture: most defaults 
occurred in 2006-2008.

```{r, defaults, echo=FALSE, message=FALSE , warning=FALSE}

# calculate default percentage by year all at once using mutate
Defaulted <- l %>%
  group_by(LoanYear) %>%
  summarise(CountTotal=n(), 
            DefaultTotal = sum(LoanStatus=="Defaulted")) %>%
  mutate(DefaultedPercent = DefaultTotal/CountTotal)

# plot of default rate by year
ggplot(data=Defaulted, aes(x=LoanYear, y=DefaultedPercent, group=1)) +
  geom_line()

```

I created the variable DefaultedPercent (# of defaulted loans / # total loans) 
to show default rates over each year. Prosper default rates seems to have 
decreased dramatically over the years. 

```{r, delinq, echo=FALSE, message=FALSE , warning=FALSE}
# only delinquent loans, labelled as past due
delin <- l[grep("Past Due", l$LoanStatus),] 

# count and sum delinquent loans
delinplot <- delin %>%
  group_by(LoanYear) %>%
  summarise(CountTotal=n(), 
            DelinTotal = sum(LoanOriginalAmount))

# plot of delinquencies by year
ggplot(data=delinplot, aes(x=LoanYear, y=CountTotal, group=1)) +
  geom_line()
```

I subsetted only deliquent loans (loans that were labelled Past Due). It turns
out there were 2067 of them. They peaked and plateaued from 2012 to 2013.

```{r, States, echo=FALSE, message=FALSE , warning=FALSE}
topstates <- l  %>% group_by(BorrowerState)  %>% summarise(count=n()) %>%
  arrange(desc(count)) %>% top_n(5)
    
ggplot(data=topstates, aes(x=BorrowerState, y=count)) +geom_bar(stat='identity')
```

I selected the top 5 states. California dominates this category.


```{r, Jobs, echo=FALSE, message=FALSE , warning=FALSE}
topoccupation <- l  %>% group_by(Occupation)  %>% summarise(count=n()) %>%
  arrange(desc(count)) %>% top_n(5)
ggplot(data=topoccupation, aes(x=Occupation, y=count))+geom_bar(stat='identity')
```

Top 5 Occupations but data seems incomplete. 



```{r, employment homeowner, echo=FALSE, message=FALSE , warning=FALSE}


ggplot(data=subset(l, l$EmploymentStatus!=""), 
       aes(x=EmploymentStatus, fill=EmploymentStatus)) + geom_bar() + 
  scale_x_discrete(labels="")


print('Is the borrower a homeowner? ')
summary(l$IsBorrowerHomeowner)
```

Borrowers are mostly employed and about half are homeowners.


# Univariate Analysis

### What is the structure of your dataset?

There are 113,937 loans in the dataset. I choose with 20 features out of 81, 
listed below. 10 being continuous variables and 10 being discrete variables 
or factors.

Continuous Variables:

EmploymentStatusDuration
CreditScoreRangeLower
CreditScoreRangeUpper
CurrentCreditLines
DebtToIncomeRatio
StatedMonthlyIncome
BorrowerAPR
BorrowerRate
LoanOriginalAmount
MonthlyLoanPayment

Discrete Variables:

BorrowerState
Occupation
EmploymentStatus
IsBorrowerHomeowner
IncomeRange
IncomeVerifiable
Term
ListingCategory
LoanStatus
LoanOriginationDate

Most loans have terms of 3 years and are under $10k (with the median 
being 6500). Most have monthly payments of usually < $500 (median of $217) 
and paying the median interest rate of 18%.
Out of 17 categories of loans, the majority of the loans are classified under 
debt consolidation (51%)
Borrowers tend to have credit scores in the IQR of 660-720 
(using creditscorerangelower). The borrower median monthly income is 4667, 
28% earns $25-50k and 27% earns $50-75K. 88% are employed in some way 
(employed, full-time or self-employed).


### What is/are the main feature(s) of interest in your dataset?

There are a number of features that will give color to the profile of the 
loan and borrower.
I think the main ones for borrowers are credit score, debt to income ratio, 
income and the ones for the loans are loan amount, interest rate, and term.


### What other features in the dataset do you think will help support your 
### investigation into your feature(s) of interest?

Loan category and status can help determine how the loans are being used
and what loans are current, defaulted, delinquent etc.

### Did you create any new variables from existing variables in the dataset?

I created TotalMonthlyDebt to estimate the total monthly debt at inception 
of loan by multiplying the DI ratio by the stated monthly income. The 
distribution is positively skewed.
I broke out the loan origination date in LoanYear, LoanMonth using lubridate. 
This really helped with yearly analysis and seasonality. 
I also created DefaultedPercent to show the default rates over each year.


### Of the features you investigated, were there any unusual distributions? 
### Did you perform any operations on the data to tidy, adjust, or change 
### the form of the data? If so, why did you do this?

I manually set two numeric categorical variables as a factor so when I plotted 
them it would look discrete: Term and ListingCategory..numeric.

# Bivariate Plots Section
```{r, correlation matrix , echo=FALSE, message=FALSE , warning=FALSE}
# Define color scheme for correlation matrix plot

col1 <- colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "white", 
                           "cyan", "#007FFF", "blue", "#00007F"))
col2 <- colorRampPalette(c("#67001F", "#B2182B", "#D6604D", "#F4A582", 
                           "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE", 
                           "#4393C3", "#2166AC", "#053061"))
col3 <- colorRampPalette(c("red", "white", "blue"))
col4 <- colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "#7FFF7F", 
    "cyan", "#007FFF", "blue", "#00007F"))
#wb <- c("white", "black")

df<- data.frame((sapply(l,c)))

# dealing with missing values
# stackoverflow.com/questions/22000630/corrplot-machinedouble-eps-missing-value
M <-cor(df, use="pairwise.complete.obs")
p = M
p[is.na(M)]=0.2 
p[is.na(M)==F]=0
M[is.na(M)]=0
corrplot(M, order = "hclust", addrect = 2, col = col4(10))


```

Correlation matrix of all variables discrete and continuous. The size of the 
dots represent the magnitude of the correlation. The colors represent
the magnitude and direction as indicated by the legend. 

All the dark blue dots (strong positive correlation), are obvious relationships.
Some variables are even derived from other. There were missing
values that I had to adjust for. The correlations are representative.
In next few scatterplots will explore the relevant strong relationships in
greater depth. 

```{r,  cl vs total month debt, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=CurrentCreditLines, y=TotalMonthlyDebt)) +
  geom_point(alpha=0.1, size= 1.5, position='jitter') +ylim(0, 7000)+
  xlim(0, 30)
```

This shows a positive correlation R^2 of .47. You have more debt as you
increase the # of credit lines or vice-versa.  

```{r,  credit vs interest rate, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=CreditScoreRangeLower, y=BorrowerRate)) +
  geom_point(alpha=0.1, size= 1, position='jitter') +xlim (450, 900) 
```

Negative correlation R^2 of -.46. Makes sense the interest rate is lower for
more creditworthy borrowers.


```{r, income vs total debt amt, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=StatedMonthlyIncome, y=TotalMonthlyDebt )) +
  geom_point(alpha=0.1, size=1.5, position='jitter') + xlim(0, 15000) +
  ylim(0, 7000) 
```

Monthly income and total monthly debt has a positive relationship of R^2 of .36. 

```{r,  loan amt vs interest rate, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=LoanOriginalAmount, y=BorrowerRate)) +
  geom_point(alpha=0.1, size= 1, position='jitter') 
```

Slight negative correlation R^2 of -.33. Seems counterintuitive, 
lower interest rate on larger loans. 


```{r,  loan amt vs credit score, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=LoanOriginalAmount, y=CreditScoreRangeLower)) +
  geom_point(alpha=0.1, size= 1, position='jitter') 
```

Slight positive correlation R^2 of -.33. The larger the loan, the higher the
credit score.


```{r,   income v d/i ratio, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=StatedMonthlyIncome, y=TotalMonthlyDebt)) +
  geom_point(alpha=0.05, size=1.5, position='jitter') + xlim(0, 15000) +
  ylim(0, 7000) + facet_wrap(~IsBorrowerHomeowner)
```

I plotted total monthly debt vs monthly income facet wrapped by 
IsBorrowerHomeowner. The homeowners have a higher dispersion whereas 
the non-homeowners are more concentrated under $5000 income. 

### A closer look at the spike in BorrowerRate 

```{r, closer look, echo=FALSE,warning=FALSE, fig.width = 8, fig.height = 10}


ggplot(data=l, aes(x=BorrowerRate, fill=ListingCategory..numeric.)) + 
  geom_bar() + xlim(0.3, .375) + 
  scale_fill_discrete(name="Loan Category", labels=cats)
```
```{r, closer look2, echo=FALSE, message=FALSE , warning=FALSE}

# create a function to interpolate between red and blue colors
# http://stackoverflow.com/questions/16295440/r-ggplot-better-gradient-color
mypal <- colorRampPalette( brewer.pal( 6 , "RdBu" ) )

#mypal(18) asks for 9 colors between red and blue for fill
ggplot(data=l, aes(x=BorrowerRate, fill=LoanYear)) + geom_bar(binwidth=0.01) + 
    xlim(0.3, .375) + scale_fill_manual(values=mypal(9))


```

Taking a closer look at spike in the histogram of BorrowerRate, these high 
interest loans were made mainly for debt consolidation in 2011-2012.

```{r, closer look 2, echo=FALSE, message=FALSE, fig.width = 8, fig.height = 10}

#mypal(18) asks for 20 colors between red and blue for fill
#http://stackoverflow.com/questions/16295440/r-ggplot-better-gradient-color

ggplot(data=subset(l, l$CreditScoreRangeLower!=0), 
       aes(x=BorrowerRate, fill=factor(CreditScoreRangeLower))) + 
  geom_bar(binwidth=0.01) + xlim(0.3, .375) + 
  scale_fill_manual("Credit Score Ranges", values=mypal(20)) 

```

And the credit score ranges are shown, the higher the borrower rate (cost)
the worse the credit rating. 


```{r,   loan amt and loan year, echo=FALSE, message=FALSE , warning=FALSE}
#by(l$LoanOriginalAmount, l$LoanYear,summary)
ggplot(data=l, aes(x=LoanYear,y=LoanOriginalAmount))+ geom_boxplot()
```

Mean loans amounts dips in 2009 and increased in 2013.

```{r,  agg loan amt and loan year, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=LoanYear, y=LoanOriginalAmount)) +geom_bar(stat='identity')
```

For comparison, aggregate loans amounts dips in 2009 and increases sharply 
in 2013, from $150M to $300M.

```{r,  loan amt over date by cat, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=l, aes(x=ldate, y=LoanOriginalAmount)) +geom_bar(stat='identity') +
  xlab("")
```

Investigating further, I graphed every loan by date, it shows a gap in service 
between November 2008 and July 2009. After doing some research on Wikipedia, 
there was a cease and desist from the SEC regarding the P2P lending model. 
The data also only goes until March 2014.


```{r,  loan amt by month, echo=FALSE, message=FALSE , warning=FALSE }
ggplot(data=l, aes(x=LoanMonth,y=LoanOriginalAmount))+ geom_boxplot()
```

Shows the seasonality of increased average loan amounts and variance during 
the holiday seasons and a pull back during the spring and summer months.


```{r, category vs loan amt, echo=FALSE,fig.width = 8, fig.height = 10,}
ggplot(data=l, aes(x=ListingCategory..numeric.,y=LoanOriginalAmount, 
                   fill=ListingCategory..numeric.,)) + 
  geom_boxplot() + scale_fill_discrete(name="Loan Category", labels=cats)
```

1. Debt consolidation and 8. Baby/Adoption are the category with the top 
amounts. ~$9k on average. 


```{r,  emp status and loan amount, echo=FALSE, message=FALSE , warning=FALSE }
ggplot(data=subset(l, l$EmploymentStatus!=""), 
       aes(x=EmploymentStatus, y=LoanOriginalAmount, color=EmploymentStatus)) +
  geom_boxplot() + scale_x_discrete(labels="")
```

Employed and fulltime on average take out larger loans.


```{r,  inc loan amount, echo=FALSE, message=FALSE , warning=FALSE }
ggplot(data=l, aes(x=IncomeRange,y=LoanOriginalAmount, color=IncomeRange))+
  geom_boxplot() + scale_x_discrete(labels="")
```

Larger incomes take out larger loans on average.


```{r,  term vs loan amt, echo=FALSE, message=FALSE , warning=FALSE }

ggplot(data=l, aes(x=Term,y=LoanOriginalAmount))+ geom_boxplot()
```

As the term is longer, the loan amounts are larger.

```{r,  term vs interest rate, echo=FALSE, message=FALSE , warning=FALSE }

ggplot(data=l, aes(x=Term,y=BorrowerRate))+ geom_boxplot()
```

But not necessarily more expensive. The mean interest rate moves up 
slightly at the 5 year term and the variance is smaller.

```{r,  term vs credit score, echo=FALSE, message=FALSE , warning=FALSE }
ggplot(data=l, aes(x=Term,y=CreditScoreRangeLower))+ geom_boxplot() +
  ylim(450, 900)
```

5 year borrowers seem to be more credit-worthy on average and also 
the variance is smaller. 


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### investigation. How did the feature(s) of interest vary with other 
### features in the dataset?
- There's a strong positive relationship between open credit lines and total
monthly debt.

- There's a negative relationship between interest rates and loan amount, the
larger the loan, the lower the rate on average. That was mostly due to them
having higher credit scores. 

- Monthly income and monthly debt have a positive relationship and levels off
after a certain point.

- Average loan amounts and variance are seasonal (higher during holidays)

- People who are employed and fulltime on average take out larger loans than
other groups.

- There is a strong positive relationship between term and loan amount
(longer the term, the larger the loan) 


### Did you observe any interesting relationships between the other features
(not the main feature(s) of interest)?

I observed that in 2008-2009 there was a large dip in loan origination.
After further investigation, it was related to a cease and desist order 
by the SEC regarding the nature of the P2P lending model. But after they 
opened again, there has been impressive growth topping out at $300 million
in loans in 2013 and close to $150 million by March of 2014 (partial year)

### What was the strongest relationship you found?

The strongest relationship I found was between credit score and interest 
rate, with R^2 .46. This makes sense since credit score is a rating of 
the credit-worthiness of the borrower and that should be directly related
to the cost of borrowing (interest rate)
Also, between open credit lines and total monthly deb with R^2 of .47. 

# Multivariate Plots Section

```{r, Borrowers segmentation, echo=FALSE, message=FALSE , warning=FALSE}
ggplot(data=subset(l, l$CreditScoreRangeLower > 660), 
       aes(x=LoanOriginalAmount, y=BorrowerRate, color=CreditScoreRangeLower)) +
  geom_point(alpha=0.5, position='jitter') + 
  scale_colour_gradient(low="blue", high="red") + 
  ggtitle("Loan Amount by Credit Score and Interest Rate")
```

The borrowers with high credit scores are on the bottom right. They generally
have lower interest rates and larger loan amounts. I subsetted only credit 
scores from 660 (1st quartile) and above for better visual presentation. 


```{r, loan normalized by credit by inc, echo=FALSE, warning=FALSE}


ggplot(data=l, aes(y=LoanOriginalAmount*CreditScoreRangeLower/850, 
                   x=IncomeRange, fill=IncomeRange )) + geom_boxplot() +
  scale_x_discrete(labels="")
```

I normalized the loan amounts by multiplying by their credit score and 
dividing by 850 (the best achievable score). The results do not differ to 
much from just using the loan amounts. 



```{r,  debt coverage , echo=FALSE, warning=FALSE,fig.width = 12, fig.height = 7}
ggplot(data=subset(l, l$IncomeRange != '$0' & l$IncomeRange != 'Not displayed'), 
       aes(x=LoanYear, y= MonthlyLoanPayment/StatedMonthlyIncome, 
           fill=IncomeRange)) +geom_boxplot() +ylim(0, 0.1) +
  facet_wrap(~IncomeRange) + theme(axis.text.x = element_text(angle=45, 
                                                              vjust=1, size=12))
```

In this plot, I divided MonthlyLoanPayment by StatedMonthlyIncome see what
the debt coverage ratio would be by year and faceted by IncomeRange. It seems
most of the borrowers with lower income ranges take on larger loans and all 
income ranges have an uptick in the most recent years.


```{r,  di ratio , echo=FALSE, warning=FALSE, fig.width = 12, fig.height = 7}

ggplot(data=subset(l, l$IncomeRange != '$0' & l$IncomeRange != 'Not displayed'), 
       aes(x=LoanYear, y= DebtToIncomeRatio, fill=IncomeRange)) +
  geom_boxplot() +ylim(0, .5) +facet_wrap(~IncomeRange)+ 
  theme(axis.text.x = element_text(angle=45, vjust=1, size=12))

```
Comparing DI ratio, most of the borrowers seem to have DI ratio close to
20-30% with a uptick in the most recent years. The $100k + income range
have noticeably lower DI ratio at around 15-20%. The variance is reduced
as incomes increase. 


```{r, loan amt over year by income, echo=FALSE, message=FALSE , warning=FALSE}
byyear <- l %>%
  group_by(LoanYear, IncomeRange) %>%
    summarise(DollarTotal=sum(LoanOriginalAmount), CountTotal=n()) %>%
  arrange(desc(DollarTotal, CountTotal))

ggplot(data=subset(byyear, byyear$IncomeRange != '$0'), 
       aes(x=LoanYear, y=DollarTotal, fill=IncomeRange)) + 
  geom_bar(stat='identity') + scale_fill_discrete(name="Income Range") +
  ggtitle("Loan Amount by Year and Income Range") + facet_wrap(~IncomeRange)

```

Looking at Income Ranges it looks like the uptick in loans in 2012-2013
grew primarily in $50-75k, $75-100k and $100k+. 


```{r,loan amt by yrcat, echo=FALSE, fig.width = 8, fig.height = 10, warning=FALSE}
byyear <- l %>%
  group_by(LoanYear, ListingCategory..numeric., IncomeRange) %>%
    summarise(DollarTotal=sum(LoanOriginalAmount), CountTotal=n()) %>%
  arrange(desc(DollarTotal, CountTotal))


ggplot(data=byyear, aes(x=LoanYear, y=DollarTotal, 
                        fill=ListingCategory..numeric.)) +
  geom_bar(stat='identity') + 
  scale_fill_discrete(name="Loan Category", labels=cats) +
  ggtitle("Loan Amount by Credit Score and Interest Rate")
```

I bucketed categories by year and you can clearly see that it looks like 
pre-2008 they did not capture the loan category and debt consolidation
dominated the cateogory every year subsequently.





# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. Were there features that strengthened each other in terms
### of looking at your feature(s) of interest?

When looking at the loan amounts vs cost (interest rate), the credit scores 
demarcated borrowers by credit worthiness. I looked at loan amounts
(normalized by credit score)  vs income ranges and saw that the higher
the income, the larger the loans amounts on average. 

### Were there any interesting or surprising interactions between features?

Investigating further, I looked at DI ratio and debt coverage ratio and
observed the higher the income, the lower the percentage of debt. 
Additionally, it appears the platform after a few years of maturity
is beginning to lend to slight more indebted borrowers in the income
ranges of $50-75k, $75-100k and $100k+. 
Looking at the debt categories, it seemed like they didn't collect this
information before 2007. But the recent uptick of lending 
is mostly in debt consolidation.


------

# Final Plots and Summary

### Plot One
```{r, Loan distribution Final, echo=FALSE, fig.width = 10, fig.height = 10}


ggplot(data=subset(l, l$CreditScoreRangeLower > 660),aes(x=LoanOriginalAmount))+
  geom_bar(binwidth=500,alpha = 0.8, color = 'blue', fill = 'lightblue') +
  facet_wrap(~LoanYear) +  ggtitle("Loan Amount Distribution by Year") + 
  xlab("Loan Original Amount ($)") + ylab("Number of loans") 
```


### Description One
This graph shows the distribution of loan amounts by year. There is a 
significant increase in loans from 2012-2013 espcially in even numbers
( $5,000, $10,000, $15,000 etc).



### Plot Two
```{r,Borrowers seg Final, echo=FALSE, fig.width = 10, fig.height = 10}

ggplot(data=subset(l, l$CreditScoreRangeLower > 660), 
       aes(x=BorrowerRate, y=LoanOriginalAmount, color=CreditScoreRangeLower))+
  geom_point(alpha=0.1, position='jitter') +
  scale_colour_gradient("Credit Score Range", low="blue", high="red") + 
  xlab("Borrower Rate") + ylab("Loan Original Amount ($)") + 
  facet_wrap(~LoanYear)+ggtitle("Loan Amount by Credit Score and Interest Rate")
```


### Description Two
The borrowers with high credit scores are in red on the left. They generally
have lower interest rates and larger loan amounts. I subsetted only credit 
scores from 660 (1st quartile) and above for better visual presentation. And 
this shows that as the lending platform matured, the overal risk exposure 
increased. In 2014, much more blue (credit score ~700) borrowers. 



### Plot Three
```{r,credit score by inc Final, echo=FALSE, fig.width = 10, fig.height = 10}
delin <- l[grep("Past Due", l$LoanStatus),] # only delinquent


ggplot(data=subset(delin, delin$IncomeRange != '$0'), 
       aes(y=LoanYear, x=LoanOriginalAmount/10000)) +
  geom_point(color='darkred', size=3, alpha=0.3, position='jitter') +
  facet_wrap(~IncomeRange) +ggtitle("Delinquent loan exposure summary") +
  xlab("Loan Original Amount $'0,000") + ylab("Loan Year")

```


### Description Three
To expand on my delinquent loans count from before. I wanted to look at
delinquent exposures across income ranges by year. It appears the majority
of Prosper's delinquencies are clustered in 2012-2013, most borrowers have
income ranges in $25k-50k and $50-75k and loan amounts under $5k.


------

# Reflection

The data set had nearly 114,000 loans from Nov 2005 - March 2014. Over the
course of those years,  Prosper has made almost $1 trillion dollars in loans
($949,894,347 to be exact). 

The difficulties I had with the data mainly stemmed from understanding the 
variables and then selecting the appropriate ones to analyze. I had to do an
initial exploration of the data and even then I probably selected a bit too 
many. I initally wanted to create a logistic regression model to see if it
could predict how likely the loan will be delinquent or even default. I would
feed the model loan stats (loan size, loan category) and borrower stats
(debt level, credit score) to see if they are good predictors. 

From the data I selected, I saw them struggle initially with their first few
years of loans. However, their business grew very quickly after the cease and 
desist by the SEC in 2009. Their main loan exposures are in debt consolidation
and I'm sure the low interest rate environment has helped the supply of
lenders looking for yield. This was a great learning experience about the P2P 
lending model and I'm glad Udacity gave me the opportunity access to this data.


